---
title: "Drawing an interactive biplot with echarts4r"
author: "Sean van der Merwe"
date: '2025-11-15'
slug: biplot-echart
categories:
  - R
  - plot
tags:
  - R
  - plot
  - interactive
code-fold: show
---

# Motivation

[Biplots](https://en.wikipedia.org/wiki/Biplot) can contain a lot of information. To dig into that information it would help if one could interact with the plot. Thus, we want to plot the information in a modern Javascript plotting standard.

# Static biplot

Let us begin by illustrating the classic iris data set with [biplotEZ](https://cran.r-project.org/package=biplotEZ).


```{r}
# Normally one might hide the setup R chunk, but here I am deliberately showing
# that I'm using SVG for static plots but best online quality
library(knitr)
fig_size <- 600
opts_chunk$set(dev = 'svg', fig.ext = 'svg', echo = TRUE, 
               fig.width = fig_size/96, fig.height = fig_size/96)
```

```{r}
#| warning: false
library(biplotEZ) |> suppressPackageStartupMessages()
```

```{r}
iris <- iris |> setNames(names(iris) |> stringr::str_replace_all("\\.", " "))
bp <- biplot(iris) |> 
  PCA(group.aes = iris$Species) |> 
  axes(col = "black") |> 
  samples(col = c("pink","orange","firebrick3"), opacity = 0.7) |>
  plot()
```

Colourful and clear 😊, but not interactive.

# Interactive version

Let's load [echarts4r](https://cran.r-project.org/package=echarts4r) and try to plot the information using it. See [this nice video introduction for more information](https://youtu.be/OBJxIWEFHdo?si=WWe5m4j4xLN3STPw).

```{r}
#| warning: false
library(echarts4r) |> suppressPackageStartupMessages()
```

The first step is to prepare the data correctly. In this case we will use the output from *biplotEZ* as a starting point.

## Points

```{r}
# Creating the point data frame is easy:
d <- data.frame(x = bp$Z[,1], y = bp$Z[,2], Species = iris$Species)
# As is plotting it:
d |> group_by(Species) |> 
  e_charts(x = x, width = fig_size, height = fig_size) |>
  e_scatter(serie = y, symbol_size = 5) |> 
  e_toolbox_feature("dataZoom") |> 
  e_tooltip(axisPointer = list(
      type = "cross"
    ))
```

**Try using the zoom tool to zoom in.**

## Axes

Those fancy axes are not so easy. Let's reconstruct them in a new way:

```{r}
create_ax_df <- function(bp) {
  xu <- bp$ax.one.unit[,1]
  yu <- bp$ax.one.unit[,2]
  tkz <- ceiling(bp$axes$ticks/2)
  ax_df <- data.frame(
    X = seq_along(xu) |> sapply(\(i) {
      seq(-tkz[i]*abs(xu[i]), tkz[i]*abs(xu[i]), abs(xu[i]))*sign(xu[i])
    }) |> c(),
    Y = seq_along(yu) |> sapply(\(i) {
      seq(-tkz[i]*abs(yu[i]), tkz[i]*abs(yu[i]), abs(yu[i]))*sign(yu[i])
    }) |> c(),
    Variable = rep(bp$axes$names, tkz*2+1),
    VarNum = rep(seq_along(bp$axes$names), tkz*2+1)
  )
  ProjMat <- as.matrix(ax_df[, 1:2]) %*% t(bp$Lmat[, 1:2])
  ax_df$Projection <- seq_len(nrow(ax_df)) |> sapply(\(i) {
      ProjMat[i, ax_df$VarNum[i]] * bp$sd[ax_df$VarNum[i]] + bp$means[ax_df$VarNum[i]]
    }) |> round(1)
  ax_df
}
v <- bp |> create_ax_df() |> group_by(Variable)
```

Then we draw a plot of the axes:

```{r}
v |>
  e_charts(x = X, width = fig_size, height = fig_size) |>
  e_line(serie = Y) |> 
  e_scatter(serie = Y)
```

## Combined

We can combine it with the scatter plot:

```{r}
d |> group_by(Species) |> 
  e_charts(x = x, width = fig_size, height = fig_size) |>
  e_scatter(serie = y, symbol_size = 5) |> 
  e_data(v, x = X) |>
  e_line(serie = Y) |> 
  e_scatter(serie = Y) |> 
  e_toolbox_feature("dataZoom") |> 
  e_tooltip(axisPointer = list(
      type = "cross"
    ))
```

The only problem is that our tooltips don't convey any useful information 😢.

## Tooltips

To have nice tooltips we first create them as another dataset variable in advance, then **bind** them to the observations as follows:

```{r}
create_tooltip <- function(df, digits = 3) {
  nms <- names(df)
  d <- seq_len(ncol(df)) |> lapply(\(j) {
    if (is.numeric(df[[j]])) {x <- df[[j]] |> round(digits = digits)} else {x <- df[[j]]}
    paste0(nms[j], ": ", x)
  }) |> list2DF()
  seq_len(nrow(d)) |> sapply(\(i) {
    paste0(d[i,] |> unlist(), collapse = "<br>")
  })
}
d$tooltip <- iris |> create_tooltip()
```

```{r}
d |> group_by(Species) |> 
  e_charts(x = x, width = fig_size*1.25, height = fig_size) |>
  e_scatter(serie = y, bind = tooltip, symbol_size = 5) |> 
  e_data(v, x = X) |>
  e_line(serie = Y) |> 
  e_scatter(serie = Y, bind = Projection, 
            label = list(show = TRUE, formatter = "{b}", offset = c(10,10))) |> 
  e_toolbox_feature("dataZoom") |> 
  e_tooltip(
    formatter = htmlwidgets::JS("
        function(params) {
          return params.name;
        }
      ")
    ) |>
  e_grid(width = "67%", right = "20%") |> 
  e_legend(
    orient = "vertical",   # vertical layout
    right = "1%",         # position from the right
    top = "middle"         # vertically centered
  )
```


# Final plot

Let's combine the new code into a function that can be added to the biplotEZ chain.

```{r}
e_biplotEZ <- function(bp, width = 600, height = 500) {
  library(echarts4r)
  d <- data.frame(x = bp$Z[,1], y = bp$Z[,2], g = bp$Xcat[[1]]) |> group_by(g)
  create_ax_df <- function(bp) {
    xu <- bp$ax.one.unit[,1]
    yu <- bp$ax.one.unit[,2]
    tkz <- ceiling(rep(5, length(xu))/2)
    ax_nms <- bp$means |> names()
    ax_df <- data.frame(
      X = seq_along(xu) |> sapply(\(i) {
        seq(-tkz[i]*abs(xu[i]), tkz[i]*abs(xu[i]), abs(xu[i]))*sign(xu[i])
      }) |> c(),
      Y = seq_along(yu) |> sapply(\(i) {
        seq(-tkz[i]*abs(yu[i]), tkz[i]*abs(yu[i]), abs(yu[i]))*sign(yu[i])
      }) |> c(),
      Variable = rep(ax_nms, tkz*2+1),
      VarNum = rep(seq_along(ax_nms), tkz*2+1)
    )
    ProjMat <- as.matrix(ax_df[, 1:2]) %*% t(bp$Lmat[, 1:2])
    ax_df$Projection <- seq_len(nrow(ax_df)) |> sapply(\(i) {
        ProjMat[i, ax_df$VarNum[i]] * (bp$sd[ax_df$VarNum[i]]^bp$scaled) + 
        (bp$means[ax_df$VarNum[i]]*bp$center)
      }) |> round(1)
    ax_df
  }
  v <- bp |> create_ax_df() |> group_by(Variable)
  create_tooltip <- function(df, digits = 3) {
    nms <- names(df)
    d <- seq_len(ncol(df)) |> lapply(\(j) {
      if (is.numeric(df[[j]])) {x <- df[[j]] |> round(digits = digits)} else {x <- df[[j]]}
      paste0(nms[j], ": ", x)
    }) |> list2DF()
    seq_len(nrow(d)) |> sapply(\(i) {
      paste0(d[i,] |> unlist(), collapse = "<br>")
    })
  }
  d$tooltip <- bp$raw.X |> create_tooltip()
  d |> 
  e_charts(x = x, width = width*1.25, height = height) |>
  e_scatter(serie = y, bind = tooltip, symbol_size = 5) |> 
  e_data(v, x = X) |>
  e_line(serie = Y) |> 
  e_scatter(serie = Y, bind = Projection, 
            label = list(show = TRUE, formatter = "{b}", offset = c(10,10))) |> 
  e_toolbox_feature("dataZoom") |> 
  e_tooltip(
    formatter = htmlwidgets::JS("
        function(params) {
          return params.name;
        }
      ")
    ) |>
  e_grid(width = "67%", right = "20%") |> 
  e_legend(
    orient = "vertical",   # vertical layout
    right = "1%",         # position from the right
    top = "middle"         # vertically centered
  )
}
```

And draw the plot clean:

```{r}
iris <- iris |> setNames(names(iris) |> stringr::str_replace_all("\\.", " "))
iris |> biplotEZ::biplot(scale = TRUE) |> 
  biplotEZ::PCA(group.aes = iris$Species) |>
  e_biplotEZ()
```








